---
title: "Project proposal"
author: "Sengdao & Alice"
output: html_document
---


```{r load-packages, message = FALSE}
library(tidyverse)
library(broom)
```

```{r}
California_housing <- read_csv("../data/California_housing .csv")
```


```{r}
glimpse(California_housing)
```

## 1. Introduction

California is one of the most volatile housing markets in the US. Housing prices have reached record highs and then crashed, leaving many residents and homeowners overwhelmed by their mortgages. This market volatility has made it difficult for first-time home buyers to enter the market and has caused many families to move out of the state in search of a more stable housing market. 

In this project, we will analyze data on the California housing market to understand what has caused the market to be volatile. We will examine data on home prices, population, total room, location, households and their income, migration patterns and many more to try to identify trends and factors that have contributed to the market instability. We hope that this analysis will provide some insight into what policy changes, if any, could be made to help stabilize the California housing market. 

## 2. Data

The California housing data set was derived from Kaggle website. This data set has 20640 obsersation and 10 variables. This data set includes information on population, total number of bedrooms, the price of the homes, and the location of the home. 

## 3. Data analysis plan

In our project, we intend to present data via table, plot, and map, and we also intend to cover codes such as summary, nrow, ncol, and so on to show the dataset's information. The goal of our project is find the relationship between the variables, and visualize it.

