---
title: "Heart disease Analysis project"
author: "Sengdao_Oudomsihn"
date: "4/15/2022"
output: html_document
---

The dataset for this statistical analysis was retrieved from the UCI Machine Learning Repository. The dataset for this statistical analysis consists of 303 observations and 14 variables. However, column names for the dataset were not included when the data was imported from the UCI Machine Learning Repository website. Below are columns names I give to the columns according to the UCI Machine Learning Repository website

Age = age in year
Sex = gender of patients; 1=male, 0=female 
CP_Type = chest pain type (Typical angina, Atypical angina, Non-anginal pain, Asymptomatic)
BloodPres = blood pressure (millimeters of mercury (mmHg))
Cholesterol = serum cholesterol in mg/dl
 Fasting_BP = fasting blood sugar >120 mg/dl (1=true; 0=false)
Rest_ECG = resting electrocardiographic results (0=normal, 1= can range from mild symptoms to severe problems, 2: possible or definite left ventricular hypertrophy
MaxHR = maximum heart rate achieved
Exercise = Exercise induced angina (1= yes, 0=no) 
ST_Dep = ST Depression induced by exercise relative to rest 
Slope_ST =The slope of the peak exercise ST segment (1:upsloping, 2: flat sloping and 3: downsloping
Num_Vsel = number of major vessels (0-3) 
Thallium_Stress =  Thallium stress result (3: normal, 6: fixed defect, 7: reversible defect
HD = Heart disease conditions


this statistical analysis aims not to classify the type of heart condition, but on the other hand, the aim of the paper is to investigate whether;
1. Age has no effect on the heart disease condition of patients
2. ST-Segment Depression has no effects the ST-elevation of heart rate 
3. Age and sex are not predictors for heart disease 
4. ST-segment Depression (ST_Dep) and ST-elevation of heart rate are predictors for different kinds of chest pains
5. There is no correlation between age and cholesterol
6. There is no correlation between blood pressure and cholesterol
7. There is no correlation between age and maximum heart rate

A few types of statistical analysis models will be used to find the answer for hypotheses. These types of statistical analysis models include one-way ANOVA, two-way ANOVA, and linear regression. Each statistical analysis model is unique in its own way, and each is used differently based on the type of data that are being compared. So the hypotheses for this analysis include many types of data. Therefore I will use different statistical analysis models for each hypothesis testing.




# tidyverse package
```{r}
# install.packages("tidyverse")
library(tidyverse)
```

# referenece source of dataset: https://archive.ics.uci.edu/ml/datasets/Heart+Disease 
```{r}
urlfile="https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data"
my_data <- read.csv(urlfile)
str(my_data)
```
# Assign names to the column names

```{r}
colnames(my_data) <- c("Age","Sex","CP_Type", "BloodPres", "Cholesterol", "Fasting_BP",  "Rest_ECG", "MaxHR", "Exercise",  "ST_Dep", "Slope_ST", "Num_Vsels", "Thallium_Stress", "HD")
str(my_data)
```

# Checking if there are  any missing values in the dataset
```{r}
## Data cleaning
any(is.na(my_data)) # check if there is any missing value in the dataset. TRUE mean yes there are missing value in the data frame
my_data[my_data == "?"] <- NA # Replace "?" in any column with NA
colSums(is.na(my_data))# which column has missing value
```

# delete anny missing value that could potentially cause problem for analysis
```{r}
my_data <- na.omit(my_data) # omit the missing value in the dataset and save it back to the dateset
any(is.na(my_data)) # False means no missing value in the dataset
```



```{r}
my_data[my_data$CP_Type == 1,]$CP_Type <- "Typical angina"
my_data[my_data$CP_Type == 2,]$CP_Type <- "Atypical angina"
my_data[my_data$CP_Type == 3,]$CP_Type <- "Non-anginal pain"
my_data[my_data$CP_Type == 4,]$CP_Type <- "Asymptomatic"
my_data[my_data$Exercise  == 0,]$Exercise  <- "No_Exercise"
my_data[my_data$Exercise  == 1,]$Exercise  <- "Yes_Exercise"
my_data[my_data$Slope_ST == 1,]$Slope_ST <- "Upsloping"
my_data[my_data$Slope_ST == 2,]$Slope_ST <- "Flatsloping"
my_data[my_data$Slope_ST == 3,]$Slope_ST <- "Downsloping"
my_data[my_data$Sex == 1,]$Sex <- "Male"
my_data[my_data$Sex == 0,]$Sex <- "Female"
my_data[my_data$HD == 0,]$HD <- "Healthy"
my_data[my_data$HD == 1,]$HD <- "Unhealthy"
my_data[my_data$HD == 2,]$HD <- "Unhealthy"
my_data[my_data$HD == 3,]$HD <- "Unhealthy"
my_data[my_data$HD == 4,]$HD <- "Unhealthy"
str(my_data)
```

# convert some variables in the dataset into appropiate data type for analysis
```{r}
# Now let's convert some variables into factor
my_data$Age <- as.integer(my_data$Age)
my_data$Sex <- as.factor(my_data$Sex)
my_data$CP_Type <- as.factor(my_data$CP_Type)
my_data$Fasting_BP <- as.factor(my_data$Fasting_BP)
my_data$Rest_ECG <- as.factor(my_data$Rest_ECG)
my_data$Exercise <- as.factor(my_data$Exercise)
my_data$Slope_ST <- as.factor(my_data$Slope_ST)
my_data$Num_Vsels <- as.factor(my_data$Num_Vsels)
my_data$Thallium_Stress <- as.factor(my_data$Thallium_Stress)
my_data$BloodPres <- as.numeric(my_data$BloodPres)
my_data$Cholesterol <- as.numeric(my_data$Cholesterol)
my_data$MaxHR <- as.numeric(my_data$MaxHR)
my_data$HD <- as.factor(my_data$HD)
str(my_data)
```
# Exploratory Data Analysis

### Histogram to see distribution of data in the dataset

```{r}
set.seed(123)
par(mfrow=c(2,3))
hist(my_data$MaxHR, breaks = 20, col="skyblue", main="Heart Rate Maximum", xlab = "Heart Rate")
hist(my_data$BloodPres,breaks = 20, col="skyblue", main="Blood Pressure", xlab = "Blood Pressure")
hist(my_data$Cholesterol, breaks = 20,col="skyblue", main = "Cholesterol Level", xlab = "Cholesterol")
hist(my_data$Age, breaks = 20, col="skyblue", main = "Age in year", xlab = "Age")
hist(my_data$ST_Dep, breaks = 20, col="skyblue", main = "ST Depression", xlab = "ST Depressoin")
```
#### Figure1: Histogram of distribution of data for maximum heart rate, blood pressure, cholesterol level, age, and ST depression for patients in the dataset. 



# density of data distribution in chest pain

```{r}
ggplot(my_data,aes(x=MaxHR, fill=CP_Type))+geom_density(alpha=.5)
```


```{r}
ggplot(data = my_data, aes(x=MaxHR, color=CP_Type, fill= CP_Type))+
  geom_histogram(aes(y=..density..), alpha=0.5,
                 position = "identity")+ 
  geom_density(alpha=.3)+
  facet_grid(CP_Type~.)+
  theme_bw()
```


```{r}
ggplot(data=my_data)+geom_bar(mapping = aes(x=Sex, fill=HD))
```


```{r}
my_data %>% group_by(CP_Type) %>% 
  count() %>% 
  arrange(desc(n))
```

```{r}
my_data %>% 
  filter(CP_Type=="Asymptomatic") %>% 
  group_by(HD, Sex, CP_Type) %>% 
  summarise(avg_HR=mean(MaxHR),
            median_HR=median(MaxHR),
            avg_age=mean(Age),
            median_age=median(Age)
            )
```

```{r}
ggplot(data=my_data)+geom_bar(mapping = aes(x=Sex, fill=HD))+facet_grid(~CP_Type)
```
```{r}
my_data %>% 
  filter(CP_Type=="Asymptomatic") %>% 
  group_by(HD, Sex, CP_Type) %>% 
  summarise_if(is.numeric, median)
```



```{r}
ggplot(data = my_data, aes(x= Sex, y= MaxHR, fill= HD))+geom_boxplot()+
  facet_grid(.~CP_Type)
```


# Statistical Analysis | Finding insight in data

## ONE-Way ANOVA

```{r}
# One-Way ANOVA
boxplot(Age~HD, data=my_data, col='tomato')
ONE_ANOVA1 <- aov(Age~HD, data=my_data)
summary(ONE_ANOVA1)
TukeyHSD(ONE_ANOVA1)
```

## Older age patients have higer unhealthy heart condition than younger age patients
Report: "Age of patients are significantly different for heart disease condition(One-way anova, F_1,294=16,52, p<0.0000617)"


```{r}
set.seed(11)
boxplot(ST_Dep~Slope_ST, data=my_data)
ONE_ANOVA2 <- aov(ST_Dep~Slope_ST, data = my_data)
ONE_ANOVA2
summary(ONE_ANOVA2)
```
## Report: "ST Depression differed significantly among the three heart rate slope(One-way ANOVA, F_2,293 =73.73, P>0.0000000000000002 or 2e-16)"

## Two-Way ANOVA

```{r}
# two-way anova
boxplot(Age~HD+Sex, data = my_data)
boxplot(ST_Dep~CP_Type+Slope_ST, data = my_data)
```


```{r}
set.seed(12)
TWO_ANOVA1 <- aov(Age~HD*Sex, data = my_data)
TWO_ANOVA1
summary(TWO_ANOVA1)
```
## Report: A two-way ANOVA analysis showed that heart condition(HD) was significantly affected by age of patients(F_1,292=16.963, p< 0.00051) and sex of patients(F_1,292=8.800, p<0.00326), with no significant different(F_1,292=0.106, p=0.74552)






```{r}
set.seed(13)
TWO_ANOVA2 <- aov(ST_Dep~CP_Type*Slope_ST, data = my_data)
TWO_ANOVA2
summary(TWO_ANOVA2)
```
## Report: A two-way ANOVA analysis showed that chest pain type(CP_Type) was significantly affected by ST Depression of patients(F_3,284=20.312, p<0.000000000592 or 5.92e-12) and slope of heart rate(Slope_ST) (F_2,284=65.750, p<0.0000000000000002 or 2e-16), with a significant different(F_6,284=4.984, p<0.00014)



```{r}
# Chest pain type as prediction for heart disease
interaction.plot(x.factor=my_data$Sex, trace.factor = my_data$HD,my_data$Age, type="o",pch=16)

# Slope of heart rate as prediction
interaction.plot(x.factor=my_data$Slope_ST,trace.factor = my_data$CP_Type,my_data$ST_Dep, type="o",pch=16)

```


## Linear Regression | Finding trends and correlation in data

```{r}
set.seed(14)
linear_model1 <- lm(data = my_data, Cholesterol~Age)
summary(linear_model1)
cor.test(my_data$Age, my_data$Cholesterol)
```

## Report: There was a significant relationship between age and cholesterol level of patients(linear regression, r^2=0.04155, F_1,294=12.75, p=0.0004165). The correlations between these two variables was positive(r=0.2038462)



```{r}
# Create linear model prediction for the confidence interval band
prediction <- predict.lm(linear_model1, data=my_data, interval = 'confidence',level=0.98)
prediction <- data.frame(prediction)
prediction$Age <- my_data$Age
prediction
```



```{r}
plot(Cholesterol~Age, data = my_data)
abline(lm(Cholesterol~Age, data = my_data))
lines(prediction$Age, prediction$lwr, col="red",lty=2)
lines(prediction$Age, prediction$upr, col="red",lty=2)
```


```{r}
ggplot(data = my_data)+geom_point(mapping = aes(x=Cholesterol, y=BloodPres, color=CP_Type, shape=CP_Type))
```





```{r}
my_data %>% ggplot(aes(x=Cholesterol, y=BloodPres))+geom_point()+
  stat_smooth(method = "lm")

# Linear Regression Model
linear_model2 <- lm(data = my_data, BloodPres~Cholesterol)
summary(linear_model2)
# run correlation
cor.test(my_data$BloodPres, my_data$Cholesterol)
```
## Report: There was a significant relationship between cholesterol level and blood pressure(linear regression, r^2=0.01752, F_1,294=5.244, p=0.02273). The correlation between these two varibales was positive(r=0.1323796)



```{r}

my_data %>% ggplot(aes(x=Age, y=MaxHR))+geom_point()+
  stat_smooth(method = "lm")

# Linear model for maximum heart rate and cholesterol
linear_model3 <- lm(data = my_data, MaxHR~Age)
summary(linear_model3)
# run correlation test
cor.test(my_data$MaxHR, my_data$Age)
```
## Report: There was a statistically significant relationship age and maximum heart rate(linear regression, r^2=0.1562, F_1,294=54.42, p=1.664e-12). The correlation between these two variables was negative(r=-0.3952039)


# Machine Learning Model

```{r}
library(caret)
library(randomForest)
library(xgboost)
```


```{r}
# data partition/ split data into training and testing
set.seed(123)
data_index <- createDataPartition(my_data$HD, p = .8, list = FALSE)
data_train <-my_data[data_index,]
data_test <- my_data[-data_index,]

```

## Random Forest machine learning model

```{r}

rf_model <- train(HD ~.,
                    data= data_test,
                    method='rf',
                    metric='Accuracy',
                    trControl=trainControl(method = 'cv', number = 10))


```


```{r}
rf_model
```




```{r}
rf_preds <- predict(rf_model, data_test)
table(rf_preds)
```


```{r}
plot(rf_preds, col='orange')
```


```{r}
plot(varImp(rf_model, scale = TRUE))
```

## Maximum heart rate is the best indicator for predicting heart condition of patients in this dataset as it shown in this visualization.  

```{r}
confusionMatrix(rf_preds, as.factor(data_test$HD))
```




